\input fontmac
\input mathmac

\font\ninebf=cmbx9

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Id{\mathop {\rm Id}}
\def\Ev{\mathop {\rm Ev}}
\def\Sym{\mathop {\rm Sym}}
\def\Aut{\mathop {\rm Aut}}
\def\Gal{\mathop {\rm Gal}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\sep{{\rm sep}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}
\def\F{{\bf F}}
\def\bsep#1{[#1]_{\sep}}

\maketitle{MATH 457 Honours Algebra 4\footnote{$^*$}{\eightpt Course given by Prof.\ Mika\"el Pichot at McGill University}}{Notes by}{Marcel K. Goh}{23 April 2020}

\floattext5 \noindent {\ninebf Note.} \ninept These notes form a large subset of the content covered by the lectures/notes given over the semester. \ninept Some proofs are either skipped or distilled to their main ideas. \ninept Note that because of the coronavirus outbreak, some concepts that are normally covered in Algebra 4 were omitted this year, e.g.\ solvability by radicals. \ninept I take responsibility for any errors that may be present.

\section 1. Rings

A {\it ring} $R$ is a set with operations $+$ and $\cdot$ such that
\medskip
\item{i)} $(R,+)$ is an abelian group;
\smallskip
\item{ii)} $(R,\cdot)$ is a semigroup;
\smallskip
\item{iii)} $\cdot$ distributes over $+$ on both sides:
$$a\cdot (b+c) = a\cdot b + a\cdot c \quad\hbox{and}\quad (a+b)\cdot c = a\cdot c + b\cdot c$$
A {\it semiring is the same as a ring} except that condition (i) above becomes
\medskip
\item{i')} $(R,+)$ is a monoid with absorbing identity 0.
\medskip
A ring is {\it unital} if $(R,\cdot)$ has a unit $1$. We always assume that $1\neq 0$, since if $1=0$ then $R = \{0\}$. Observe that in a unital ring, $(R,+)$ is necessarily abelian. A ring is said to be {\it commutative} if $(R,\cdot)$ is.

Even for commutative rings, there are many possible ring structures for $(R,+) = \Z^2$. For example we can take the {\it Gaussian integers} $\Z[i] = \{a+bi : a,b\in\Z\}$ or the {\it Eisenstein integers} $\Z[\omega] = \{a + b\omega : a,b\in \Z\}$ where
$$\omega = -{1 + i\sqrt{3} \over 2}.$$
In both cases the second binary operation is complex multiplication. Since $i$ and $\omega$ are both solutions to equations of the form $x^2 + Bx + C = 0$, they are called {\it quadratic integers} and $\Z[i]$ and $\Z[\omega]$ are called {\it quadratic rings}.

The definition of a ring is meant to describe a class of $\Z$-like objects, but many rings have properties different from the integers. For example, the ring $\Z[\sqrt{-5}]$ does not have Euclidean division. There are also many non-commutative rings such as the {\it Lipschitz quaternions}
$$\{a + bi + cj + dk : a,b,c,d\in \Z\}$$
or the {\it Hurwitz quaternions}
$$\bigg\{a + bi + cj + dk : a,b,c,d\in \Z\ \hbox{or}\ a,b,c,d\in \Z + {1\over 2}\bigg\}.$$

If $R$ is a ring, then a subgroup of $(R,+)$ that is closed under multiplication is called a {\it subring}. If a ring is unital, then any unital subring will have the same unit. A {\it homomorphism} between two rings $R$ and $S$ is a map $f:R\to S$ that preserves both operations:
$$f(a+b) = f(a) + f(b) \qquad\hbox{and}\qquad f(a\cdot b) = f(a)\cdot f(b)$$
A homomorphism that preserves the units is called {\it unital}.

An {\it ideal} in a ring $R$ is a subgroup $(I,+)$ such that
\medskip
\item{i)} $ab\in I$ for all $a\in R$, $b\in I$;
\smallskip
\item{ii)} $ab\in I$ for all $b\in I$, $a\in R$.
\medskip
If (i) holds, $I$ is called a {\it left ideal} and if (ii) holds, $I$ is called a right ideal. Let $I \subseteq R$ be an ideal. One defines the {\it quotient ring} $R/I$ as follows. Since $(I,+)$ is a normal subgroup of $(R,+)$, $R/I$ is an abelian group. We associate $r\sim r'$ if $r-r'\in I$. Then we can define multiplication in $R/I$ as $(a+I)(b+I) = ab+I$. This is well-defined because $I$ is an ideal and distributivity holds.

The isomorphism theorems for groups extend to rings as well. Their proofs are analogous to the ones in group theory.

\parenproclaim Theorem A (First isomorphism theorem). Let $f:R\surj S$ be a surjective ring homomorphism. Then $f$ descends to a a ring homomorphism $f' : R/I \to S$ that takes $a+I$ to $f(a)$, where $I$ is the kernel of $f$.\slug

\parenproclaim Theorem B (Second isomorphism theorem). Let $S$ be a subring and $I$ and ideal in a ring $R$. Then $S+I$ is a subring of $R$, $I$ is an ideal in $S+I$, and the map $S \surj (S+I)/I$ is a surjective ring homomorphism with kernel $S\cap I$.\slug

\parenproclaim Theorem C (Third isomorphism theorem). Let $R$ be a ring and $I\subseteq J\subseteq R$ be ideals. Then $R/I \surj R/J$ is a surjective ring homomorphism with kernel $J/I$.\slug

\parenproclaim Theorem D (Fourth isomorphism theorem). Let $f:R\surj S$ be a surjective ring homomorphism. There is a bijection between the ideals in $R$ containing $\ker f$ and the set of all ideals in $S$.\slug

Note that the correspondence in Theorem D works with subrings as well, not just ideals.

An element $r$ in a unital ring $R$ is said to be {\it invertible} if there exists $s\in R$ such that $rs = sr = 1$. The set of invertible elements is denoted $R^\times$ and this is a group under $\times$, called the {\it group of units}. A {\it field} is a ring in which every nonzero element is a unit. Non-commutative fields are called {\it division rings} or {\it skew fields} (the quaternions are an example of a skew field).

Let $K$ be a field. The set $K[x]$ of polynomials with coefficients in $K$ is a ring. Then the set
$$K(x) = \{ f / g : f,g\in K[x], g\neq 0\}$$
is a field, called the {\it field of rational functions}. The set $K[[x]]$ is called the {\it ring of formal series}: possibly infinite sums $\sum_{n\geq 0} a_nx^n$. Addition is done pointwise and multiplication is convolution of power series. The map $K[x] \to K[[x]]$ is a homomorphism and some elements become invertible. For example, $1-x$ becomes invertible, since $1/(1-x) = \sum_{n\geq 0} x^n$. Not every element in $K[[x]]$ is invertible, but one can invert the elements to get a new field $K((x))$: the set of sequences $K^\Z$ that are eventually zero when going to the left.

A {\it zero-divisor} is an element $r\in R$, $r\neq 0$ for which there exists $s\in R$ such that $rs = 0$. A ring is {\it cancellative} if $rs = rs'$ implies that $s = s'$. Then we define an {\it integral domain} to be a unital, commutative, cancellative ring. Every integral domain embeds into a field, called the {\it field of fractions}. The construction is analogous to building the rational numbers from the integers.

\proclaim Proposition Z. If $R$ is a ring with unity, there exists a unique unital homomorphism $f:\Z\to R$.\slug

\proof Since $f(1) = 1$, we have $f(n) = 1 + 1 + \cdots + 1 \in R$.\slug

The nonnegative integer $n$ which generates $\ker f$ is called the {\it characteristic} of $R$. The image of $f$ is called the {\it characteristic subring}. For example $\Z/n\Z$ has characteristic $n$.

\proclaim Proposition P. The characteristic of an integral domain $R$ is either $0$ or a prime number.\slug

An {\it algebra} over a commutative ring $R$ is a ring $A$ with a homomorphism $\eta : \R\to A$ whose image lies in the {\it centre} of $A$. Examples of algebras include rings of functions and matrices $M_n(R)$.

For a group $G$ and a ring $R$, we can define the {\it group ring} $G[R]$ as the set of all finitely supported functions from $G$ to $R$. This forms a ring with addition $(f+g)(s) = f(s) = g(s)$ and multiplication $(fg)(s) = \sum_{uv=s} f(u)g(u)$.

\section 2. Ideals

Every element $r$ in a unital ring $R$ generates a {\it principal} ideal $(r)$. More generally any subset $S\subseteq R$ does. The ideal $(S)$ is the intersection of all ideals that contain $S$. If $R$ is commutative, then $(r) = rR = Rr$. In $\Z$, the ideals are the of the form $(n) = n\Z$. Then $(n)\subseteq (m)$ if and only if $m\setminus n$ (this is true in any commutative ring). A ring $R$ in which every ideal is principal is called a {\it principal ring} and if $R$ is also an integral domain, we call it a {\it principal ideal domain} or PID.

Principal ideals determine their generators up to unit. If $(r) = (s)$, then $s = ar$ and $r = bs$ together imply that both $a$ and $b$ are units. Elements $r$ and $s$ of a ring $R$ are called {\it associate} if there exists a unit $a$ such that $r=as$.

We can define three operations on ideals. Let $I,J\subseteq R$ be ideals.
\medskip
\item{i)} $I\cap J$ is an ideal.
\smallskip
\item{ii)} $I+J = \{a + b : a\in I, b\in J\} = (I\cup J)$ is an ideal.
\smallskip
\item{iii)} $IJ = \{ab : a\in I, b\in J\}$ is an ideal.
\medskip

\proclaim Lemma P. Let $R$ be a commutative ring. Let $I = (S)$ and $J = (T)$ be two ideals. Then $IJ = (ST)$. \slug

In the ring of integers $\Z$, we have $(m)(n) = (mn)$, $(m)\cap(n) = \big(\lcm(m,n)\big)$, and $(m)+(n) = \big(\gcd(m,n)\big)$. When $I\subseteq J$ is an inclusion of ideals, one may think of it as a kind of divisibility $J\setminus I$. For example, $\gcd(m,n)\setminus \lcm(m,n)\setminus mn$.

\proclaim Lemma D. If $I,J\subseteq R$ are ideals, then
$$IJ \subseteq I\cap J \subseteq I+J.\noskipslug$$

The set of ideals forms a semiring where the two operations are $I+J$ and $IJ$. The semiring in $\Z$ is $\N$ with the addition $m + n = \gcd(m,n)$ and ordinary multiplication.

For an ideal $I\subseteq R$, we define the {\it radical} of $I$ to be the set
$$\sqrt{I} = \{a \in R : a^n \in I\ \hbox{for some}\ n\in \N\}.$$
This is an ideal and it has the property that $\sqrt{\sqrt{I}} = \sqrt{I}$. Furthermore, if $I\subseteq J$, then $\sqrt{I} \subseteq \sqrt{J}$.

An ideal $I\subseteq R$ is called {\it maximal} if it is proper and whenever $I\subseteq J\subseteq R$, then either $J = I$ or $J=R$.

\proclaim Lemma M. Let $R$ be a unital ring. Then every proper ideal is included in a maximal ideal.

\proof This is an application of Zorn's Lemma. Let $I$ be a proper ideal and let $X$ be the set of all proper ideals containing $I$, ordered by inclusion. Then this set is inductive (increasing union of ideals is an ideal) so there is a maximal element $M$.\slug

\proclaim Lemma F. Let $R$ be unital and commutative. Then an ideal $I\subseteq R$ is maximal if and only if $R/I$ is a field.

\proof This follows from the fourth isomorphism theorem.\slug

Let $R$ be a unital ring. An ideal $I$ of $R$ is {\it prime} if it is proper and for any ideals $A,B$ of $R$, $AB\subseteq I$ implies that $A\subseteq I$ or $B\subseteq I$. The {\it spectrum} of $R$ is the set of all prime ideals and it is denoted $\Spec(R)$. The {\it maximal spectrum} of $R$, denoted $\Specmax(R)$, is the set of all maximal ideals of $R$.

Maximal ideals are always prime (so $\Specmax(R) \subseteq \Spec(R)$), but not all prime ideals are maximal. For example, $(0)$ is prime in $\Z$ but certainly not maximal. A ring is called {\it local} if it has a unique maximal ideal. A ring $R$ is local if and only if $R\setminus R^\times$ is an ideal.

\proclaim Lemma C. Let $R$ be a unital commutative ring. Let $I\subseteq R$ be a proper ideal. Then $I$ is prime if and only if $ab\in I$ implies that $a\in I$ or $b\in $I.\slug

\proclaim Lemma I. Let $R$ be a unital commutative ring. Then $I\subseteq R$ is a prime ideal if and only if $R/I$ is an integral domain.\slug

Since all fields are integral domains, this proves that all maximal ideals are prime. We also have that a commutative ring $R$ is an integral domain if and only if $(0)$ is a prime ideal in $R$ (if $R$ is not commutative, then we say it is a {\it prime ring}). If $R$ is a PID, then every nonzero prime ideal is maximal.

We can view elements in a commutative unital ring $R$ as ``functions'' on the set $\Spec(R)$ of prime ideals. To $r\in R$ we identify a function $f_r$ such that $f_r(P) = r \bmod P\in R/P$. We have a bundle at every $P\in \Spec(R)$ and a fibre $R/P$ which is an integral domain. The {\it total space} $B(R)$ is the union of $R/P$ over all prime ideals $P$. A {\it section} is a map $s: \Spec(R)\to B(R)$ such that $s(P)\in R/P$. $\Gamma(R)$ is the set of all sections and $\Gamma_{\max}(R)$ is its restriction to $\Specmax(R)$. Let $\pi : R\to \Gamma(R)$ map $r\mapsto f_r$ and $\pi_{\max} : R\to \Gamma_{\max}(R)$ take $r$ to $f_r$, restricted to $\Specmax(R)$. We want to know when $\pi$ and $\pi_{\max}$ are faithful.

\proclaim Proposition K. The kernel of $\pi$ is the intersection of all prime ideals and the kernel of $\pi_{\max}$ is the intersection of all maximal ideals.\slug

For a unital commutative ring $R$, we define the {\it nilradical} of $R$ to be the intersection $\Nil(R) = \bigcap P$ of all prime ideals $P$. The {\it Jacobean radical} is the intersection $\Jac(R) = \bigcap M$ of all maximal ideals $M$. Since $\Specmax(R) \subseteq \Spec(R)$, $\Jac(R) \supseteq \Nil(R)$. An element $r\neq 0$ in a ring $R$ is called {\it nilpotent} if $r^n = 0$ for some $n$. It turns out that there is a connection between nilpotency and prime ideals.

\proclaim Proposition N. Let $R$ be unital and commutative. Then $\Nil(R)$ is the set of all nilpotent elements, i.e.
$$\sqrt{(0)} = \{r\in R : r^n = 0\ \hbox{for some}\ n\in \N\} = \bigcap_{P\in \Spec(R)} P.$$

\proof To show that a nilpotent element $r$ belongs to every prime ideal $P$, note that $r^n \in P$, so $r\cdot r^{n-1}\in P$ and we can iterate this until we get that $r\in P$. Conversely, if $r$ is not nilpotent, we can let $X$ be the set of ideals $I$ such that $r^n$ is not in $I$ for any $n$. $X$ is nonempty and inductive, so by Zorn's Lemma there is a maximal element and it can be shown that this ideal is prime.\slug

Let $R$ be a commutative ring and let $p\in R$ be a nonzero non-unit. Then $p$ is said to be
\medskip
\item{i)} {\it prime} if $p\setminus ab$ implies that $p\setminus a$ or $p\setminus b$;
\smallskip
\item{ii)} {\it irreducible} if $p=ab$ implies $a$ is a unit or $b$ is a unit.
\medskip

To find irreducible elements in a ring, may attempt the ``bisection process''. Let $r\in R$. If $r$ is irreducible, we stop. If $r$ is not irreducible, then $r = r_1 r_2$. If neither is irreducible, we continue by splitting $r_1$ and $r_2$ in the same way. This process may not terminate.

\proclaim Proposition I. Let $R$ be an integral domain. If an element $p\in R$ is prime, then it is irreducible.

\proof. Let $p\in R$ be a prime element. Assume that $p=ab$. This implies that $p\setminus a$ or $p\setminus b$. Say $a = pc$ for some $c\in R$. Then $p=ab=pcb$ and $cb = 1$. So $b$ is a unit.\slug

Note that the converse does not hold. For example, in the ring $\Z[\sqrt{-3}]$, we have $4=(1+\sqrt {-3})(1-\sqrt {-3})$. The element 2 is irreducible, but it is not prime because $2$ divides $4$ but does not divide either of $(1+\sqrt{-3})$ and $(1-\sqrt{-3})$.

\proclaim Proposition A. Let $R$ be an integral domain. Let $p$ be a nonzero element in $R$. Then $p$ is prime if and only if $(p)$ is prime and $p$ is irreducible if and only if $(p)$ is maximal among principal ideals.\slug

This proposition implies that in a PID, irreducible elements are prime.

A ring $R$ is a {\it unique factorisation domain} if every $r\in R$ can be expressed as a product $r = p_1\cdots p_n$ of irreducible elements, which is unique up to the order of the $p_i$. The rings $\Z$, $K[x]$, and $K[x,y]$ are all examples of UFDs. Every PID is a UFD and in a UFD, all irreducible elements are prime.

\proclaim Lemma S. In a PID, every chain of ideals stabilises.

\proof $I = \bigcup_{n\geq 1} I_n$ is an ideal. Since $R$ is a PID, $I = (x)$ for some $x$ and $x\in I_n$ for some $n$. This implies that $I = I_n$.\slug

\proclaim Lemma N. Let $R$ be a unital ring. Then every increasing chain of ideals stabilises if and only if every ideal is finitely generated.

\proof If $I = (x_1,x_2,\ldots)$ is not finitely generated, then $I_n = (x_1,\ldots,x_n)$ is an increasing chain of ideals that does not stabilise. Conversely, if every ideal is finitely generated, then let $I_1\subseteq I_2\subseteq \cdots$ be a chain of ideals and let $I = \bigcup_{n\geq 1} I_n$. There exist $(x_1,\ldots,x_n)$ that generate $I$, so there exists a $k$ such $x_i\in I_k$ for all $i$ and we find that $I = I_k$.\slug

A ring is called {\it Noetherian} if it the equivalent conditions from Lemma N hold.

For elements $r$ and $s$ of a ring, a {\it greatest common divisor} or gcd is an element $d$ dividing both $r$ and $s$ such that if any $d'$ divides both $r$ and $s$, then $d'$ divides $d$. An integral domain $R$ is called a {\it B\'ezout domain} if $(r)+(s)$ is principal for every $r,s\in R$ (of course, every PID is a B\'ezout domain) and it is called a {\it GCD domain} if any two $r,s\in R$ have a gcd. Every UFD is a GCD domain.

\proclaim Lemma B. The following statements regarding B\'ezout domains are true.
\medskip
\item{i)} A ring $R$ is B\'ezout if and only if every finitely generated ideal is principal.
\smallskip
\item{ii)} A B\'ezout domain is a GCD.
\smallskip
\item{iii)} If a ring is both Noetherian and a B\'ezout domain, then it is a PID.\noskipslug
\medskip

\section 3. Gaussian Integers

Recall from Section 1 that the Gaussian integers are the ring
$$\Z[i] = \{a + bi : a,b\in \Z\}.$$
We write $N$ for the complex modulus, squared. So $N(z) = z\bar z = a^2 + b^2$. This is called the {\it norm} and it is a group homomorphism $\C^\times\to \R^\times$, since $N(zz') = N(z)N(z')$. $N(z) = 0$ implies that $z = 0$. The norm $n$ takes $\Z[i]$ to $\N$. The kernel of $N$ on $\C^\times$ is the unit circle $\{z\in \C : |z| = 1\}$. Let $\ker N$ denote the kernel of $N$ restricted to $\Z[i]$, i.e.\ $\{\pm 1, \pm i\}$. These are the units of $\Z[i]$.

The image of $N$ is
$$\Im(N) = \{n\in \N : n = a^2 + b^2\ \hbox{for some}\ a,b\in \Z\}.$$
This set is stable under product, since if $n = N(z)$ and $n' = N(z')$, then $nn' = N(zz')$. Gauss was interested in studying the number of integer numbers less than a given $n$ that can be expressed as a sum of two squares. We will return to this point later.

We say that a prime number {\it splits} if it is no longer prime in $\Z[i]$ and we say that it is {\it inert} otherwise.

\proclaim Lemma S. Let $p$ be a prime. Then $p$ is a sum of two squares if and only if it splits in $\Z[i]$.

\proof If $p = a^2 + b^2$ then $p = (a + ib)(a-ib)$ and $N(a+ib)N(a-ib) = p^2$ implies that neither of these factors are units. So $p$ is not prime in $\Z[i]$. Conversely, if $p = \alpha\beta$ in $\Z[i]$, then $N(\alpha) = N(\beta) = p$ means that $p$ is the sum of two squares.\slug

\proclaim Lemma I. A prime $p$ splits if and only if $p\equiv 1 \mod 4$.

\proof If $p$ splits, then by the previous lemma, $p = a^2 + b^2$ and the sum of two squares is never $3$ modulo 4. So if $p$ is an odd prime it is congruent to 1 modulo 4. Conversely, assume that $p\equiv 1\mod 4$. Then $p = 1 + 4n$ for some $n$ and there exists $x\in \Z$ such that $x^2 \equiv 1\mod p$. (In fact, $x = (2n)!$ works.) Then $p$ divides $x^2 + 1 = (x+i)(x-i)$. So $p$ divides $(x+1)$ or $(x-i)$, so $p$ divides $i$ and is not inert.\slug

All this talk of divisibility leads nicely into a discussion of Euclidean division. In $\Z$, the goal of Euclidean division for integers $a$ and $b$ is to find a $q\in \Z$ such that $a-bq$ is small, in some sense. The following proves a similar result in $\Z[i]$.

\proclaim Proposition E. There is a Euclidean division in $\Z[i]$.

\proof. Let $a,b\in \Z[i]$, $b\neq 0$. We can divide them in $\C$ to get $z = a/b$. Then there is a (not necessarily unique) $q\in \Z[i]$ that is of minimal distance to $z$. We have $|z-q|<1$; in fact $|z-q| \leq \sqrt{2}/2 < 1$. So $|a-bq| < |b|$.\slug

Let us now define this generally. An integral domain $R$ is a {\it Euclidean domain} if there exists a function $N : R\to \N$ called the {\it norm} such that $N(0) = 0$ and for all $a,b\in R$, $b\neq 0$, either $b$ divides $a$ or there exists $q\in R$ such that $N(a-bq) < N(b)$. Proposition E showed that $Z[i]$ is a Euclidean domain with the complex norm, and other familiar examples include $\Z$ with the absolute value function and $K[x]$ with the degree of a polynomial as its norm. In general, the Euclidean division algorithm does not give a unique answer. Even in $\Z$, we can end up with $q$ or $-q$ as a quotient.

\proclaim Proposition T. $\Z[\sqrt{-2}]$ is a Euclidean domain.

\proof We repeat the same proof as for $\Z[i]$ except for the computation of $|z-q|$, which is now $\leq \sqrt{3}/2$.\slug

Recall that $\Z[\sqrt{-3}]$ is not a Euclidean domain. It is not even a UFD, since $4 = 2\cdot 2 = (1+\sqrt{-3})(1-\sqrt{-3})$. But $\Z[\sqrt{-3}]\subseteq \Z[\omega]$ and this is a Euclidean domain, with norm $N(a+b\omega) = a^2 - ab + b^2$. The units in $\Z[\omega]$ are the elements of norm 1: $\{\pm 1, \pm \omega, \pm \omega^2\}$ and we have unique factorisation up to units.

\proclaim Proposition P. Every Euclidean domain is a PID (and consequently a UFD).

\proof Let $R$ be a Euclidean domain and $I\subseteq R$ an ideal. Let $b\neq 0$ be an element of $I$ of minimal norm. If $a\in I$ then $b$ divides $a$. Otherwise, there exists $q\in R$ such that $N(a-bq) < N(b)$, contradicting the minimality of $b$'s norm. So $I$ is principal.\slug

A corollary of this fact is that every ideal in $\Z[i]$ is principal.

\section 4. Modules

For any set $X$, the set of symmetries $\Sym(X)$ is a group and an action of a group $G$ on $X$ is a group homomorphism $G\to \Sym(X)$. If $X$ is a group, we can define the {\it ring of endomorphisms} $\End(X)$ as the set of group homomorphisms from $X$ to $X$.

\proclaim Lemma M. Let $M$ be an abelian group. Then $\End(M)$ is a ring.

\proof Addition is pointwise addition from $M$ and multiplication is composition of maps.\slug

Let $R$ be a a unital commutative ring. A {\it module} $M$ over $R$ is a ring homomorphism $R\to \End(M)$. Explicitly, the list of axioms of a module are very similar to those of a vector space (in fact, if $R$ is a field, then a module is a vector space). For $r,s\in R$ and $m,n\in M$, we have
\medskip
\item{i)} $r(m+n) = rm + rn$;
\smallskip
\item{ii)} $(r+s)m) = rm + sm$;
\smallskip
\item{iii)} $(rs)m = r(sm)$;
\smallskip
\item{iv)} $1m = m$.
\medskip
These axioms also work if $R$ is not commutative; in this case, we call $M$ a {\it left $R$-module}. The kernel of $R\to \End(M)$ is called the {\it annihilator} of $M$:
$$\Ann(M) = \{r\in R : rm = 0\ \hbox{for all}\ m\in M\}$$
A module is said to be {\it faithful} if $\Ann(M)$ is trivial. If $M$ is an $R$-module, then $M$ is a faithful $S$-module where $S = R/\Ann(M)$.

\proclaim Proposition I. Any ideal $I$ in a ring $R$ is a module over $R$.

\proof For $a\in I$ and $r\in R$, we have $ra\in I$. The rest of the axioms follow.\slug

Quotients $R/I$ are also modules. When $R = \Z$, the action is determined by the group structure in $M$. For example,
$$2m = (1 + 1)m = 1m + 1m = m + m.$$
When $R = K[x]$ for some field $K$, we have the following interesting lemma.

\proclaim Lemma V. $K[x]$-modules are operators on vector spaces and vice versa.

\proof Let $M$ be a $K[x]$-module. The restriction of the $K[x]$ action to $K$ gives a $K$-module structure on $M$. This is a vector space. Furthermore, the indeterminate $x$ also acts on $M$ by taking $m\mapsto xm$. This gives a map $x: M \to M$ such that $x(m + n) = x(m)$ and $x(rm) = (xr)m = (rx)m = r\cdot x(m)$. So $x$ is a linear map.

Conversely, if $V$ is a $K$-vector space, and $T:V\to V$ is a linear map, then $V$ is a $K[x]$-module, because for any $p\in K[x]$, $p(T)$ is a linear map on $V$.\slug

Note that the module is not faithful, because $K[x]$ has infinite dimension but $\End(V)$ has finite dimension when $V$ has finite dimension. If $G$ is a group, $K[G]$ is the group ring and a $K[G]$-module is a linear representation of $G$.

A {\it submodule} $M'$ of $M$ is a subgroup that is stable under the action of the ring, i.e.\ for all $m, n\in M'$ and $r\in R$, $m+rn\in M'$. For example, ideals are submodules of $R$ and if $M$' is a submodule, we can define the {\it quotient module} $M/M'$ with the action of $R$:
$$ r(m+M') = rm + M'$$
If $M$ and $M'$ are modules, then $M\times M'$ is a module. If a module has no proper nontrivial submodules, then it is called {\it simple}.

An $R$-module map is a group homomorphism $f:M\to M'$ such that $f(rm) = rf(m)$ for all $r\in R$ and $m\in M$. The kernel $\ker f$ is a submodule of $M$ and the image of $f$ is a submodule of $M'$. The isomorphism theorems for modules are exactly analogous to the ones given for rings in Section 1.

\parenproclaim Lemma S (Schur's lemma). Let $M$ be a simple module. Then $\End_R(M)$ is a skew field.

\proof Let $f:M\to M'$ be a module map that is not identically zero. The kernel of $f$ is a submodule of $M$, so since $f\neq 0$, $\ker f = \{0\}$. Then the image of $f$ is a submodule of $M'$ since $f\neq 0$, $\Im f = M'$. Hence $f$ is an isomorphism.\slug

If $M$ is an $R$-module and $I\subseteq R$ is an ideal, then
$$IM = \Big\{\sum r_im_i : r_i\in I, m_i\in M\Big\} \subseteq M$$
is a submodule.

\parenproclaim Theorem C (Chinese remainder theorem). Let $I,J$ be ideals in a ring $R$. Let $M$ be an $R$-module. Then the map
$$M\to M/IM \times M/JM$$
has kernel $IM\cap JM$.\slug

If $I+J = R$ then the map is surjective and $(I\cap J)M = IJM$. With $n$ ideals such that $I_k + I_l = R$ for $k\neq l$, we have
$$M/(I_1\cdots I_n)M \cong M/I_1M \times \cdots \times M/I_nM.$$

Let $M$ be an $R$-module. If $A\subseteq M$, then
$$(A) = \Big\{\sum r_ia_i : r_i\in R, a_i \in A\Big\}$$
is the submodule of $M$ {\it generated} by $A$. A module is {\it finitely generated} if it admits a finite generating set and {\it cyclic} (or {\it singly generated}) if it is generated by one element. If $M = (a)$ is cyclic, then the map $R\to M$ that sends $r\mapsto ra$ is surjective with kernel $\Ann(M)$.

\proclaim Lemma P. Let $R$ be an integral domain. Then the nonzero principal ideals are isomorphic to $R$.

\proof Let $I = (a)$ be an ideal (so it is an $R$-module). If $r\in \Ann(I)$ then $r$ is a zero divisor. So $R\to I$ is an isomorphism.\slug

A finitely generated $R$-module $M$ is called {\it free} if it is isomorphic to $R^n$ for some $n$. For example, if $R$ is a field, every module (finite-dimensional vector space) is free. Equivalently, an $R$-module is free if there exists a basis, that is, a generating set $A$ such that any $m\in M$ can be written in a unique way as a finite sum
$$m = \sum_{a\in A} r_a a.$$
The set $A$ is called a {\it free generating set} and the cardinality of $A$ is called the {\it rank} of $M$.

In a PID, every ideal is a free module (isomorphic to the ring itself). For any set $A$ and ring $R$, we can let $F_A$ be the set of all functions from $A$ to $R$ with finite support. This is a group under pointwise addition and $r$ acts on $F_A$: $(rf)(a) = r\cdot f(a)$. A basis for $F_A$ is the set $(\delta_a)_{a\in A}$ of delta functions, where $\delta_a(b) = 1$ if $b=a$ and $0$ otherwise.

\parenproclaim Proposition U (Universal property of free modules). Let $\phi$ be a map from a set $A$ to an $R$-module $M$. Then there is a unique extension of $\phi$ to a module map $\bar\phi : F_A\to M$.

\proof Take any element $f\in F_A$ and express it as
$$f = \sum_{a\in A} r_a\delta_a$$
for some $r_a\in R$. Then let $\bar\phi$ be given by
$$\bar\phi(f) = \sum_{a\in A} r_a\phi(a).\noskipslug$$

\proclaim Proposition S. Let $N\inj M \surj F$ be a short exact sequence of modules (so $F \cong N/M$), where $F$ is a free module. Then the sequence splits, i.e.\ $M\cong N\oplus F$.

\proof We need to construct the section $s$ of $\pi : M\surj F$. Let $A$ be a basis of $F$. Since $\pi$ is surjective, for any $a\in A$ we can find $m_a\in M$ such that $\pi(m_a) = m$. This gives a map $s_* : A \to M$ and by the universal property there is a unique extension $s : F_A\to M$. We have $\pi \circ s = \Id$ on the basis and therefore everywhere on $F$. So $s$ is a section of $\pi$. Let $F' = \Im(s) \subseteq M$. So $F'\cong F$ as $R$-modules. We claim that $M = N\oplus F'$ (viewing $N$ as a submodule of $M$).

Firstly, $N\cap F' = \{0\}$, since if $m\in N\cap F'$, then $\pi(m) = 0$ and there exists $f\in F$ such that $s(f) = m$. But this implies that $f = \pi\big(s(f)\big) = \pi(m) = 0$, so $m=0$. And $M = N + F'$ because any $m\in M$ can be expressed as the sum of $\big(m-s\circ \pi(m)\big) + s\circ\pi (m)$.\slug

The following theorem shows that the rank of a free module is well-defined.

\proclaim Theorem R. If $R^n\cong R^m$ as $R$-modules, then $n = m$.

\proof Since $R$ is unital and commutative, it contains a maximal ideal $M$. Let $K = R/M$ and consider the submodule $MR^n = \big\{s(x_1,\cdots,x_n) \in R^n: s\in M, x_i\in R\big\}$. The quotient module is $K^n = (R/M)^n$ and a module isomorphism $R^n \cong R^m$ descends to a $R$-module isomorphism $K^n \cong K^m$. This map has kernel $M$ and is a $K$-vector space isomorphism. So the dimension of the two vector spaces are the same and thus $n=m$.\slug

Let $M$ be a module over an integral domain $R$. The set of {\it torsion elements}
$$\Tor(M) = \{m\in M : rm = 0\ \hbox{for some}\ r\neq 0\}$$
is a submodule of $M$. A module is called {\it torsion} if $\Tor(M) = M$ and {\it torsion-free} if $\Tor(M) = \{0\}$. Note that $R^n$ is torsion-free, since it has a basis $\{e_n\}$ and if $am = ra_1e_1 + \cdots + ra_ne_n = 0$, then $ra_i = 0$ for all $i$ and $m = 0$.

\proclaim Proposition T. For any module $M$ over an integral domain $R$, $M/Tor(M)$ is torsion-free.

\proof Let $N = M/\Tor(M)$ and let $\bar m\in N$. Suppose there exists $r\neq 0$ such that $r\bar m = 0$. So $\bar{rm} = 0$ and $rm \in \Tor(M)$. Thus there exists $s\neq 0$ such that $(rs)\bar m = 0$. But $\bar{rs}\neq 0$ so $m$ must be 0. Hence $\Tor(N) = \{0\}$.\slug

\proclaim Lemma G. A module $M$ over an integral domain $R$ is torsion if and only if it is generated by torsion elements.

\proof The forward direction is clear. Conversely, suppose $M = (A)$ and every element in $A$ is torsion. Let $m = s_1a_1 + \cdots + s_na_n \in M$ for some $s_i\in R$ and $a_i\in A$. Each $a_i$ is a torsion element so there is $r_i$ such that $r_ia_i = 0$. Let $r = r_1\cdots r_n$. Then $rm = 0$.\slug

\proclaim Proposition F. Let $M$ be a finitely generated module over an integral domain $R$. There exists a free module $F\subseteq M$ such that $M/F$ is torsion.

\proof Let $M = (A)$ where $A$ is finite. Let $B\subseteq A$ be a maximal basis which generates a free module $F$ of rank $n = |B|$. Let $N$ be the quotient $M/F$. For every $a\in A\setminus B$, the module $(B\cup \{a\})$ is not free. So there exists $r\in R, r_b\in R$, not all zero, such that
$$ra + \sum_{b\in B} r_b b = 0.$$
Note that $r\neq 0$, otherwise $B$ would not be a basis. But $ra = 0 \bmod F$, so $N$ is generated by torsion elements and by the previous lemma, $N$ is torsion.\slug

\section 5. Modules Over PIDs

An $R$-module has properties very much like a vector space when $R$ is a PID.

\proclaim Proposition F. Let $R$ be a PID. Then every submodule of a free module $R^n$ is free of rank $k\leq n$.

\proof We proceed by induction. When $n=1$, every ideal $I\subseteq R$ is free, isomorphic to $R$. Now assume the proposition is true for $R^n$. Let $M$ be a submodule of $R^{n+1}$. Let $\pi : R^{n+1} \surj R^n$ be the projection map on to the first $n$ coordinates. So we have a short exact sequence
$$\ker(\pi_{|M}) \inj M \surj \pi(M).$$
But $\pi(M)$ is a submodule of $R^n$ so, by the induction hypothesis, it is free and the sequence splits. Thus $M\cong \ker(\pi_{|M}) \oplus \pi(M)$ is free.\slug

A module $M$ over a unital ring $R$ is called a {\it Noetherian module} if every submodule is finitely generated.

\proclaim Proposition N. Let $M$ be a left $R$-module. The following are equivalent:
\medskip
\item{i)} $M$ is Noetherian.
\smallskip
\item{ii)} $M$ satisfies the ascending chain condition on left modules.
\smallskip
\item{iii)} If $\frak F$ is a nonempty family of submodules, there exists a maximal element in $\frak F$ with respect to inclusion.
\medskip

\proof To show that (i) implies (ii), we let $N_1\subseteq N_2 \subseteq N_3 \subseteq \cdots$ be an increasing sequence of modules. We need to know there is an upper bound. Let $N = \bigcup_{i\geq 1} N_i$. Since $N$ is finitely generated, there will be a first index $i$ such that all the generators of $N$ belong to $N_i$. Thus $N = N_i$ for some $i$.

We show that (ii) implies (iii) by contraposition. If (iii) fails, then there exists a family $\frak F$ of submodules for which there is no maximal element. Pick $N_1\in \frak F$. We can find $N_2$ such that $N_1$ is properly contained in $N_2$. Continuing in this way, we are left with an increasing chain that does not stabilise.

Lastly, assume that (i) does not hold; i.e.\ there is a submodule $N$ that is not finitely generated. Let $\{n_1,n_2,\ldots\}$ be an infinite countable subset of $N$ such that, for every $k$, $N_k = (n_1,\ldots,n_k)$ is properly contained in $N_{k+1} = (n_1,\ldots,n_{k+1})$. Now $\frak F = \{N_k\}$ is a family of submodules without a maximal element, so (iii) fails.\slug

\proclaim Proposition S. Let $N \inj P\surj Q$ be an exact sequence of modules. Then $N$ and $Q$ are Noetherian if and only if $P$ is Noetherian.

\proof Clearly $N$ is Noetherian if $P$ is. To show that $Q$ is Noetherian, let $M\subseteq Q$ be a submodule. Then, by the fourth isomorphism theorem, $M$ is the image of a submodule $M'$ of $P$. Since $M'$ is finitely generated, $M$ is as well.

Conversely, assume that $N$ and $Q$ are Noetherian and let $M\subseteq P$ be a submodule. Let $\pi:P\surj Q$ be the quotient map and consider the exact sequence $\ker(\pi_{|M})\inj M\surj \pi(M)$. Let $X$ be a finite generating set for $\ker(\pi_{|M})$ and $Y$ be a finite set in $M$ such that $\bar Y = \pi(Y)$ is a finite generating set of $\pi(M)$. Then for any $m\in M$, then there exist some $r_x$ and $r_y$ in $R$ such that
$$m = \sum_{x\in X} r_x x + \sum_{y\in Y} r_y y$$
and $X\cup Y$ generates $M$.\slug

\proclaim Theorem R. The following are equivalent:
\medskip
\item{i)} $R$ is a Noetherian ring.
\smallskip
\item{ii)} The free module $R^n$ is Noetherian for every $n$.
\smallskip
\item{iii)} Every finitely generated $R$-module is Noetherian.\noskipslug
\medskip

A corollary of Theorem R is that every finitely generated module over a PID is Noetherian. For example, $R = K[x_1,\ldots,x_n]$ is Noetherian.

\proclaim Lemma N. If $R$ is a PID and $M$ a torsion-free $R$-module, then $M$ is free.

\proof In general, we showed that there exists $F$ free such that $F\inj M \surj T$, where $T$ is torsion. Since $M$ is Noetherian, we can choose a maximal $F$ satisfying this property. We claim that $M=F$. Let $\pi:M\surj T$ denote the quotient map and let $m\in M$. Since $\pi(m)\in T$ is a torsion element, there exists $r\in R$ such that $r\pi(m) = 0$. So $rm\in \ker\pi$ and $rm\in F$. Let $f_r : M \to M$ be the map that sends $m\mapsto rm$. This map is injective because $M$ is torsion-free. Since $f_r(F) \subseteq F$ and $f_r(M) \subseteq F$, so the submodule $f_r(F,M)$ is contained in $F$. By Proposition F, $f_r(F,M)$ is free, so $M$ is free.\slug

\proclaim Theorem T. Let $M$ be a finitely generated module over a PID $R$. Then $M\cong R^n \oplus \Tor(M)$.

\proof Consider the exact sequence $\Tor(M)\inj M\surj N$ where $N$ is torsion-free. Since $R$ is a PID, $N$ is free and the sequence splits, giving us the desired direct sum decomposition.\slug

The integer $n$ given by Theorem T is called the {\it free rank} of a module. If two modules $M$ and $N$ are isomorphic, then their free ranks are equal and $\Tor(M)\cong \Tor(N)$. The following theorem is called the structure theorem for finitely generated modules over a PID.

\proclaim Theorem S. Let $R$ be a PID and let $F \cong R^n$ be a finitely generated free module. Let $M$ be a finitely generated submodule of $F$. Then there exists a basis $(e_1,\ldots, e_n)$ of $F$ and elements $r_1,\ldots,r_m\in R$ such that $(r_1e_1,\ldots,r_me_m)$ forms a basis of $M$:
$$F/M \cong R^{n-m} \oplus R/(r_1) \oplus \cdots \oplus R/(r_m)$$
The elements $r_i$ are unique up to multiplication by a unit if we assume that $r_i$ divides $r_{i+1}$.\slug

The elements $r_i$ in Theorem S are called the {\it invariant factors} of the module.

\section 6. Fields and Polynomials

Because the kernel of a homomorphism is an ideal, then any nontrivial homomorphism $f:K\to R$ is injective when $K$ is a field. If $R = L$ is a field, then $K\subseteq L$ is a subfield and we call $L$ an {\it extension} of $K$. We will  often denote this by $L/K$. If $L/K$ is a field extension then $L$ is a vector space over $K$ and the dimension $[L:k] = \dim_K L$ is called the {\it degree} of the extension. Every there is a basis $(\alpha_i)$ of $L$ such that any element $l\in L$ can be expressed as $\lambda_1\alpha_1 + \cdots \lambda_n\alpha_n$ where $\lambda_i\in K$. If $K\subseteq L\subseteq M$ is a chain of extensions and $(\beta_j)$ is a basis of $M$ over $L$, then it can shown that $(\alpha_i\beta_j)$ is a basis of $M$ over $K$. So $[M:K] = [M:L][L:K]$.

A field is {\it prime} if it contains no proper nontrivial subfields. Any field $K$ is an extension of a prime field that contains $1$, $1+1$, etc. as well as their inverses. If the characteristic of $K$ is 0, then the prime field is $\Q$, and if the characteristic is a prime $p$, then the prime field is $\F_p$.

\proclaim Lemma F. Let $K$ be a field of characteristic $p$. The Frobenius map $x\mapsto x^p$ is a field homomorphism (so it is injective).

\proof For any $x,y\in K$, we have $(x+y)^p = x^p + y^p$ (by the binomial theorem) and $(xy)^p = x^py^p$.\slug

Let $L/K$ be an extension and $S\subseteq L$ be a set. Then $K(S)$ is the subfield of $L$ generated by $S$. The extension field is finitely generated if $S$ is finite. If $S$ consists of a single element $\alpha$, then $L = K(\alpha)$ is called a {\it simple} extension and $\alpha$ is a {\it primitive element}. For $\alpha_1,\ldots,\alpha_n$, the extensions $K(\alpha_1,\ldots,\alpha_n)$ and $K(\alpha_1)\cdots(\alpha_n)$ are the same and the order in which the elements are adjoined does not matter.

If $K$ is a field then $K[x]$ is a PID. So for any irreducible polynomial $f\in K[x]$, $(f)$ is maximal and $L = K[x]/(f)$ is a field extension. This is called the {\it Kronecker construction}.

\proclaim Lemma R. Every $f\in K[x]$ admits a root in a finite-degree extension.

\proof We may assume that $f$ is irreducible. It is of finite degree so $L = K[x]/(f)$ is a finite degree extension and if $\alpha = x \bmod f$, then $f(\alpha) = 0$ in $L$.\slug

Kronecker's construction is universal in the following sense. Let $L/K$ be an arbitrary extension and let $\alpha\in L$. Consider the map $\Ev_\alpha : K[\alpha] \to L$ that takes a polynomial $f$ to $f(\alpha)$. Since $K[x]$ is a PID, the $\ker(\Ev_\alpha)$ is principal and equals $(f_\alpha)$ for some polynomial $f_a\in K[x]$. Because $L$ is an integral domain, one of two things may happen. The first is that $f_\alpha$ is an irreducible polynomial, in which case we say that $\alpha$ is {\it algebraic}. The second is that the kernel is trivial and in this case we call $\alpha$ {\it transcendental}.

When $\alpha$ is algebraic, the unique monic irreducible polynomial $f_\alpha$ such that $f_\alpha(\alpha) = 0$ is called the {\it minimal polynomial} of $\alpha$ and $K(\alpha) \subseteq L$ is obtained by the Kronecker construction
$$K(\alpha)\cong K[x]/(f_\alpha).$$
If $\alpha$ is transcendental, $\Ev_\alpha : K[x] \inj L$ is injective and it extends to the fraction field $K(x) \inj L$ by taking $f/g \mapsto f(\alpha)/g(\alpha)$ (since $g(\alpha) \neq 0$ whenever $g\neq 0)$. Liouville established the existence of transcendental numbers in 1844 by proving that
$$L = \sum_{n\geq 0} {1\over 10^{n!}}$$
is transcendental. We have $\Q(L)\cong \Q(x) \subseteq \R$. Other famous transcendental numbers are $\pi$ and $e$.

An extension $L/K$ is {\it algebraic} if every $\alpha\in L$ is algebraic over $K$ and the following lemma proves some properties of algebraic extensions.

\proclaim Lemma A. Assume that an extension $L$ is generated by $\alpha_1,\ldots,\alpha_n$ over $K$. The following are equivalent:
\medskip
\item{i)} The elements $\alpha_1,\ldots,\alpha_n$ are algebraic over $K$.
\smallskip
\item{ii)} The degree $[L:K]$ is finite.
\smallskip
\item{iii)} Every $\alpha\in L$ is algebraic over $K$.
\medskip

\proof Suppose (i) holds. We have
$$K \subseteq K(\alpha_1)\subseteq K(\alpha_1, \alpha_2) \subseteq \cdots \subseteq L,$$
and since each $\alpha_i$ is algebraic over $K$, it is algebraic over $K(\alpha_1,\ldots,\alpha_{i-1}$, whence
$$\big[ K(\alpha_1,\ldots,\alpha_i) : K(\alpha,\ldots,\alpha_{i-1})\big] < \infty.$$
By the multiplicativity of the degree, $[L:K] \leq \infty$.

Suppose (iii) fails, i.e.\ some $\alpha\in L$ is not algebraic. Then $K(x) \cong K(\alpha) \inj L$ is an injection into $L$, contradicting the fact that $[K(x) : K] < \infty$. Thus (ii) implies (iii).

That (iii) implies (i) is obvious, so we are done.\slug

To construct extensions of a field, we need to find irreducible polynomials. Over $\Q$, we can consider $x^n - p$ where $p$ is a prime. Then $\Q(\root n\of p)$ is an extension of degree $n$ over $Q$. A general check for irreducibility is given by the following criterion.

\parenproclaim Theorem E (Eisenstein's criterion). Let $R$ be an integral domain and let $f\in R[x]$ be a monic polynomial of degree $n$. If there exists a prime ideal $\frak p$ such that $f = x^n \bmod \frak p$ and $f(0) \notin \frak p^2$, then $f$ is irreducible.

\proof Suppose, towards a contradiction, that $f = ab$ is reducible. Then, we have $x^n = \bar a \bar b$, where the bar indicates the polynomials modulo $\frak p$. In particular, $\bar a\bar b$ has zero constant term. The ideal $\frak p$ is prime, so $R/\frak p$ is an integral domain, so both $\bar a$ and $\bar b$ have zero constant term modulo $\frak p$, meaning that the constant terms of $a$ and $b$ both belong to $\frak p$. This is a contradiction, since it is clear that the constant term of $f$ belongs to $\frak p^2$.\slug

We can use Eisenstein's criterion to show that cyclotomic polynomials of the form
$$\Phi_p(x) = {x^p -1\over x-1} = x^{p-1} + x^{p-2} + \cdots + x + 1$$
for $p$ prime are irreducible. The criterion does not immediately apply, but if we consider
$$\Phi_p(x+1) = x^{p-1} + px^{p-2} + {p(p-1)\over 2}x + p,$$
we find that Eisenstein's criterion applies, so $\Phi_p(x+1)$ is irreducible and this implies that $\Phi_p(x)$ is irreducible, since any factorisation for $\Phi_p(x)$ would give a factorisation for $\Phi_p(x+1)$ (replacing $x$ with $x+1$).

There are other many other criteria for reducibility/irreducibility; we give two more famous ones.

\parenproclaim Theorem C (Cohn's criterion). Suppose that $p$ is prime and in some base $b$,
$$p = a_nb^n + \cdots + a_1b + a_0.$$
Then $f = a_nx^n + \cdots + a_1x + a_0$ is irreducible in $\Z[x]$.\slug

\parenproclaim Lemma G (Gauss' lemma). Let $R$ be a UFD with fraction field $K$ and let $f\in R[x]$ be a polynomial of degree $n$ such that $\gcd(a_n, \ldots, a_0) = 1$. If $f$ is reducible in $K[x]$, then $f$ is reducible in $R[x]$.\slug

Gauss' lemma is often applied with $R = \Z$ and $K = \Q$. It says that it $f$ is irreducible in $\Z[x]$, then it is irreducible in $\Q[x]$.

\section 7. Splitting Fields

We begin with a lemma regarding the interchangeability of the roots of a polynomial.

\proclaim Lemma I. Let $f\in K[x]$ be a monic, irreducible polynomial and let $L/K$ be an extension of the field $K$. If $\alpha,\beta$ are two roots of $f$ in $L$, then there is a field isomorphism $K(\alpha) \cong K(\beta)$.

\proof This follows from the universality of the Kronecker construction: $K(\alpha) \cong K[x]/(f) \cong K(\beta)$.\slug

More generally, any field isomorphism $\phi : K\to L$ extends uniquely to a ring isomorphism $\bar\phi : K[x] \to L[x]$ defined by applying $\phi$ on the coefficients. Then $f\in K[x]$ is irreducible if and only if $\bar\phi(f)$ is irreducible. Let $\alpha$ be an arbitrary root of an irreducible polynomial $f\in K[x]$ and let $\beta$ be an arbitrary root of $\bar\phi(f)$. Then there exists a unique field isomorphism $\phi^* : K(\alpha) \to L(\beta)$ that takes $\alpha$ to $\beta$ and whose restriction to $K$ is $\phi$. A corollary of this fact is that if $f\in K[x]$ is irreducible, then all roots of $f$ have the same multiplicity in an algebraic closure (this will be expanded on later).

A {\it splitting field} for a polynomial $f\in K[x]$ is an extension $L/K$ such that
$$f = \prod_{i=1}^n (x - \alpha_i)$$
for $\alpha_i\in L$ and $L = K(\alpha_1,\ldots,\alpha_n)$.

\proclaim Proposition S. Every polynomial $f\in K[x]$ of degree $n$ admits a splitting field of degree at most $n!$.

\proof Let $\alpha_1$ be an abstract root of $f$ in $K(\alpha_1)$ obtained by the Kronecker construction. Then $f = (x-\alpha_1)f_1$ for some $f_1\in K(\alpha_1)$. Let $\alpha_2$ be a root of an irreducible factor of $f_1$ and extend the field to $K(\alpha_1,\alpha_2)$. This process happens at most $n$ times, by which time we will have found a splitting field $L$ of $f$. We have
$$K \subseteq K(\alpha_1) \subseteq K(\alpha_1,\alpha_2) \subseteq \cdots \subseteq L,$$
and the degree of each $f_i$ is $n-i$. So by multiplicity of the degrees we have $[L:K] \leq n!$.\slug

For example, the polynomial $f = x^4 - 1$ has roots $\pm 1, \pm i$. When $K = \Q$, the abstract bound for the degree of the splitting field is $4! = 24$, but in fact $\Q(i)$ is a splitting field for $f$, of degree 2. More generally, when $f = x^n - 1$, the roots are the $n$th roots of unity $1, \omega, \ldots, \omega^{n-1} \in\C$, where $\omega = e^{2ki\pi/n}$ for $i = 0,\ldots,n-1$. The roots form a group isomorphic to $\Z/n\Z = \langle\omega\rangle$ and $K(\omega)$ is the splitting field of $x^n - 1$, since if one root is added, all of its powers come along for the ride. If $n$ is prime, then the degree of this splitting field is $n-1$; in general, the degree is equal to $\varphi(n)$ where $\varphi$ denotes Euler's totient function.

The following theorem shows that splitting fields are unique up to $K$-isomorphism.

\parenproclaim Theorem K (Kronecker{\rm, 1887}). Let $f\in K[x]$ be an irreducible polynomial. Let $\alpha$ and $\alpha'$ be two roots of $f$ in two splitting fields $L/K$ and $L'/K$ respectively. Assume the existence of a map $\theta_0 \in \Aut(K)$ that fixes coefficients of $f$. Then there exists an isomorphism $\theta : L \to L'$ such that $\theta_{|K} = \theta_0$ and $\theta(\alpha) = \alpha'$.

\proof The proof is by strong induction on $n = \deg f$. If $f$ splits in $K[x]$, then $L = K = L'$. We can take $\theta = \theta_0$, finishing the case when $n=1$ and when every irreducible factor of $f$ has degree 1.

Now we assume that the theorem is proved for any field $K$, automorphism $\theta_0$ and polynomial $f$ of degree less than $n$. Now let $p\in K[x]$ be an irreducible factor of $f$ of degree at least 2. If $\alpha\in L$ and $\alpha'\in L'$ are roots of $p$, then we can extend $\theta_0$ to an isomorphism $\theta : K(\alpha) \to K(\alpha')$. Let $K_1 = K(\alpha)$ and ${K_1}' = K(\alpha')$ for short. We have $f = (x-\alpha)f_1$ in $K_1$ and $f = (x-\alpha'){f_1}'$ in ${K_1}'$ where $f_1$ and ${f_1}'$ have degree $n-1$. Now $L$ is a splitting field for $f_1$ over $K_1$ and $L'$ is a splitting field for ${f_1}'$ over ${K_1}'$. Since the degrees of $f_1$ and ${f_1}'$ are less than $n$, by the induction hypothesis there is an isomorphism $\theta^* : L \to L$ that extends the isomorphism $\theta : K_1 \to {K_1}'$. The restriction of $\theta^*$ to $K_1$ is $\theta$, and the restriction of that onto $K$ is $\theta_0$.\slug

Let $L/K$ and $L'/K$ be field extensions. A {\it $K$-embedding} $L \inj L$ is an injective homomorphism that fixes $K$. If $\theta : L\to L'$ is a bijection, we call it a {\it $K$-automorphism}. The {\it Galois group} of a polynomial $f\in K[x]$ is the group of $K$-automorphisms of a splitting field of $K$. If $L$ is a splitting field, we denote this group by $\Gal(L/K)$ or $\Aut(L/K)$. For short, we may use the notation $\Gal(f)$ for a polynomial $f$, but this is only well-defined up to conjugacy. If $L/K$ and $L'/K$ are two splitting fields, then by Theorem K there exists a $K$-isomorphism $\theta:L\to L'$ and $\Aut(L/K) \cong \Aut(L'/K)$.

\proclaim Lemma A. Let $R$ be the roots of a polynomial $f$. Then $\Gal(f)$ acts on $R$.

\proof If $\theta\in \Gal(f)$ and $f(\alpha) = 0$, we have
$$\alpha^n + \lambda_{n-1}\alpha^{n-1} + \cdots + \lambda_1\alpha + \lambda_0 = 0$$
and
$$\theta(\alpha)^n + \lambda_{n-1}\theta(\alpha)^{n-1} + \cdots + \lambda_1\theta(\alpha) + \lambda_0 = 0$$
so $\theta(\alpha)\in R$. This defines an action $\Gal(f) \to \Sym(R)$.\slug

\proclaim Proposition F. The action of $\Gal(f)$ on the set of roots $R$ is faithful.

\proof Let $L = K(R) = K(\alpha_1,\ldots,\alpha_n)$ be a splitting field. Suppose that $\theta\in \Gal(f)$ acts trivially on $R$, i.e.\ $\theta(\alpha_i) = \alpha_i$ for all $i$. Since $\theta_{|K} = \Id$, $\theta = \Id$ as an automorphism of $L$.\slug

We have established that $\Gal(f) \inj \Sym(R)$ is a group of permutations of the roots of $f$.

\proclaim Proposition O. Let $\Gal(f)$ act on the set $R$ of roots of $f$. Then the orbit $\Gal(f)\alpha$ of $\alpha\in R$ is the set $R_1$ of roots of $f_1$ where $f_1$ is the irreducible factor of $f$ such that $f_1(\alpha) = 0$.

\proof If $f_1(\alpha) = 0$, since $f_1\in K[x]$. we have $f_1\big(\theta(\alpha)\big) = 0$ so $\theta(\alpha)\in R_1$ (the set of roots of $f_1$). Thus $\theta(R) \subseteq R_1$. Furthermore, since $f_1$ is irreducible, Kronecker's uniqueness theorem shows that for any two roots $\alpha$ and $\beta$ of $f_1$, there exists some $\theta_i : L\to L$ such that $\theta_i(\alpha) = \beta$. So $\Gal(f)\alpha = R_1$.\slug

As a corollary, if $f$ is already irreducible, then $\Gal(f)$ is transitive. In general, it is not tractable to classify all the transitive subgroups (up to conjugacy) of $S_n$. This has been done for small $n$, however; for example, when $n = 6$ there are 16 different possible subgroups.

Let us look at an example. Consider $f = (x^2 - 2)(x^2 - 3)$ over the field $K = \Q$. Then $R = \{\pm \alpha = \sqrt 2, \pm \beta = \sqrt 3\}$ and $L = \Q(\alpha, \beta)$. The Galois group cannot take $\sqrt 2$ to $\sqrt 3$ because such a permutation does not preserve the relations between the roots: If $\theta(\alpha) = \beta$, then $2 = \theta(\alpha^2) = \theta(\alpha)^2 = 3$, a contradiction. In this case, $\Gal(f)$ is the Klein four-group $V$. Let $\theta_2$ be the permutation that switches $\pm \sqrt 2$ and let $\theta_3$ transpose $\pm \sqrt 3$. Then the two commute and generate a group isomorphic to $V$.

As another example, consider $f = x^4 - 2$ over $K = \Q$. Eisenstein's criterion with $p=2$ tells us that $f$ is irreducible, and the set of roots turns out to be $R = \{\pm \alpha = \root 4 \of 2, \pm \beta = i \root 4 \of 2\}$. The splitting field is $L = \Q(\alpha, \beta)$. We can employ a useful trick, namely that {\sl if the coefficients of $f$ are real, then complex conjugation permutes the roots.} This gives us an element in $\Gal(f)$ of order 2: the permutation that fixes $\alpha$ and takes $\beta\mapsto -\beta$. In any case, we will employ a more general method to compute $\Gal(f)$. Let $\theta\in \Gal(f)$. Since $\alpha^2 + \beta^2 = 0$, we have
$$\theta(\alpha)^2 + \theta(\beta)^2 = 0.$$
Suppose that $\theta(\alpha) = \beta$. Then $\theta(\beta)^2 = -\beta^2$ and $\theta(\beta)$ is $\pm i\beta$. If $\theta(\beta) = -i\beta = \alpha$, then we have the order 2 automorphism
$$s = (\alpha \leftrightarrow \beta, -\alpha\leftrightarrow-\beta).$$
If, instead, we have $\theta(\beta) = i\beta = -\alpha$, then we have the order 4 automorphism
$$r = (\alpha \mapsto \beta \mapsto -\alpha \mapsto -\beta \mapsto \alpha).$$
We conclude that $\Gal(f) = \langle r,s \rangle = D_4$.

A field is {\it algebraically closed} if every $f\in K[x]$ admits a root in $K$. For example, $\C$ is algebraically closed. Every algebraically closed field is infinite. The Kronecker construction tells us that for any finite set $F\subseteq K[x]$, there is a finite extension $L$ of $K$ such that every polynomial $f\in F$ splits in $L$. We can extend our definition of a splitting field to (not necessarily finite) subsets of $K[x]$. Then an {\it algebraic closure} of $F$ is a splitting field for $K[x]$.

The following theorem gives the existence and uniqueness (up to $K$-isomorphism) of the algebraic closure $\bar K$ for every field $K$. The construction is ``universal'' in the sense that if $L/K$ is an algebraic extension, then there exists a $K$-embedding $L\inj \bar K$.

\parenproclaim Theorem S (Steinitz{\rm, 1910}). Let $K$ be a field. There exists an algebraic closure $\bar K$ of $K$ and this extension is unique up to $K$-isomorphism.

\proof First we show existence. Let $\frak A$ be the set of all algebraic extensions of $K$, ordered by set inclusion. Observe that $\frak A$ is inductive; indeed, if $L_1\subseteq L_2\subseteq \cdots$ is a chain, then $\bigcup_{i=1}^\infty L_i$ is in $\frak A$. By Zorn's Lemma, there exists a maximal element of $\frak A$, call it $L$. We claim that $L$ is algebraically closed. Let $f\in L[x]$ be a polynomial. By the Kronecker construction, there is an extension $L(\alpha)$ of $L$. Then $L(\alpha)$ is algebraic and $L\subseteq L(\alpha)$. Since $L$ is a maximal element in $\frak A$, $L = L(\alpha)$ and $f$ admits a root in $L$.

Next we show uniqueness of the algebraic closure. It is enough to show that if $L$ is algebraic over $K$, then $L\inj \bar K$ (if there were two algebraic closures, this would make them isomorphic). So fix an algebraic extension $L/K$. Consider the set of all intermediate extensions $K\subseteq L_\phi\subseteq L$ and for each $L_\phi$ there exists an embedding $\phi : L_\phi \to \bar K$. Let $\frak B$ be the set of all such $\phi$, partially ordered in the following manner: $\phi \leq \phi'$ if $L_\phi \subseteq L_{\phi'}$, i.e.\ $\phi'$ extends $\phi$.

The poset $\frak B$ is inductive, since if $\phi_1 \leq \phi_2 \leq \cdots$, then we can let
$$L_\phi = \bigcup_{i=1}^\infty L_{\phi_i}.$$
For any $\alpha\in L_\phi$, we have $\alpha\in L_{\phi_i}$ for some $i$ and we can simply set $\phi(\alpha) = \phi_i(\alpha)$. (This does not depend on the choice of $i$ since the functions extend one another.) By Zorn's Lemma, $\frak B$ admits a maximal element $\phi : L_\phi \to \bar K$. We want to show that $L_\phi = L$. If not, then there exists $\alpha\in L\setminus L_\phi$. By the uniqueness of Kronecker, $\phi: L_\phi \to \bar K$ admits an extension $\bar\phi : L_\phi(\alpha) \to \bar K$. But since $\bar \phi$ is an extension of $\phi$, by the maximality of $\phi$ we have $\phi= \bar\phi$ so $\alpha\in L_\phi$.\slug

As corollaries of Theorem S we have $\bar K = \bar{\bar K}$ and if $L/\bar K$ is algebraic, then $L = \bar K$.

\proclaim Proposition R. Let $K$ be a field and $f\in K[x]$ monic and irreducible. Let $\alpha,\beta\in \bar K$ be roots of $f$. Then there exists $\theta\in \Aut_K(\bar K)$ such that $\theta(\alpha) = \beta$. Conversely, if $\theta(\alpha) = \beta$ then $\alpha$ and $\beta$ have the same minimal polynomial.

\proof Theorem K provides an isomorphism $\theta : K(\alpha) \to K(\beta)$ such that $\theta(\alpha) = \beta$. Then Theorem S extends $\theta$ (not uniquely) to a map in $\Aut_K(\bar K)$. Conversely, if $\theta$ fixes $K$ then it fixes the coefficients of the minimal polynomial of $\alpha$ and $\beta$.\slug

The group $\Aut_K(\bar K)$ is called the {\it absolute Galois group} over $K$ and it acts on $\bar K$ with finite orbits. The orbits are precisely the set of roots of irreducible polynomials in $K[x]$.

\proclaim Theorem N. Let $K \subseteq L\subseteq \bar K$ be an intermediate extension of $K$. The following are equivalent:
\medskip
\item{i)} $L/K$ is a splitting field for a subset of polynomials over $K$.
\smallskip
\item{ii)} Every irreducible polynomial $f\in K[x]$ which admits a root in $L$ splits in $L$.
\smallskip
\item{iii)} For all $\theta\in \Aut_K(\bar K)$, $\theta(L) = L$.
\medskip

\proof Conditions (i) and (ii) are clearly equivalent. To show that (ii) implies (iii), let $\theta\in \Aut_K(\bar K)$ and $\alpha\in L$. There exists $f\in K[x]$ irreducible such that $f(\alpha) = 0$. So $\theta(\alpha)$ is a root of $f$ and $\theta(\alpha)\in L$. Lastly, suppose (iii) holds. Let $f\in K[x]$ be an irreducible polynomial and let $\alpha\in L$ be a root of $f$. If $\beta$ is another root of $f$, by Proposition R there is a $\theta\in \Aut_K(\bar K)$ that takes $\alpha$ to $\beta$, and by hypothesis, $\theta(L) = L$, so $\beta\in L$.\slug

Any intermediate extension satisfying the equivalent conditions of Theorem N is called a {\it normal} extension. For any normal extension $L$, and any $\theta\in \Aut_K(\bar K)$, we can restrict $\theta$ to $L$. This gives a surjective group homomorphism from $\Aut_K(\bar K)$ to $\Aut_K(L)$. The latter is denoted $\Gal(L/K)$ and is called the {\it Galois group} of $L/K$.

\section 8. Separable Extensions

Besides normality, there is another important concept in the Galois theory called separability. Let $f\in K[x]$ be irreducible and $L = K[x]/(f) = K(\alpha)$ be the Kronecker extension, where $\alpha$ is the abstract root, and let $\bar K$ be an algebraic closure of $K$. How many $K$-embeddings $L\inj \bar K$ are there? This number is called the {\it separable degree} of $L/K$ and is denoted $\bsep{L:K}$. This is the degree of an intermediate subfield $L_\sep\subseteq L$, called the {\it separable closure} of $K$ in $L$. Then
$$\bsep{L:K} = [L_\sep : K].$$
We define the {\it separable degree} $\deg_\sep(f)$ to be the number fo distinct roots of $f$ in a splitting field. Thus if $\alpha$ is a root in an extension $K(\alpha)$ of $K$ and $f_\alpha$ is its minimal polynomial, then
$$\bsep{K(\alpha):K} = \deg_\sep(f_\alpha).$$
A polynomial $f$ is called {\it separable} if its roots are all distinct, i.e.\ $\deg(f) = \deg_\sep(f)$. We can make similar definitions for an element $\alpha$ of an extension. Its {\it separable degree} is the separable degree of its minimal polynomial and it is {\it separable} if its minimal polynomial is separable. This is equivalent to $\bsep{K(\alpha):K} = [K(\alpha):K]$.
An algebraic extension is {\it separable} if it contains only separable elements. The ``universal property'' in Theorem 7S gives us the multiplicativity formula
$$\bsep{M:K} = \bsep{M:L} \bsep{L:K}$$
whenever $K\subseteq L\subseteq M$ is a chain of algebraic extensions.

\proclaim Proposition M. Let $L = K(\alpha_1,\alpha_2,\ldots,\alpha_n)$ be an algebraic extension. Then $\bsep{L:K} \leq [L:K]$ and the following are equivalent:
\medskip
\item{i)} The elements $\alpha_1,\alpha_2,\ldots,\alpha_n$ are separable over $K$.
\smallskip
\item{ii)} $\bsep{L:K} = [L:K]$.
\smallskip
\item{iii)} Every element of $L$ is separable over $K$.
\medskip

\proof By invoking the multiplicativity of degrees, it is enough to prove the proposition for a primitive extension $K(\alpha)$. The equivalence of (i) and (ii) is discussed above and (iii) trivially implies (i). So we prove that (ii) implies (iii). If $\beta\in K(\alpha)$, then $K\subseteq K(\beta) \subseteq K(\alpha) = L$ and
$$[L:K] = [K(\alpha) : K(\beta)][K(\beta) : K].$$
If $\alpha$ is separable over $K$, then it is also separable over $K(\beta)$ so
$$[K(\alpha):K(\beta)] = \bsep{K(\alpha):K(\beta)}.$$
Thus if we assume that $\bsep{L:K} = [L:K]$, then we also have $\bsep{K(\beta) : K} = [K(\beta) : K]$ so $\beta$ is separable over $K$.\slug

A {\it Galois extension} is an extension that is algebraic, normal, and separable.

\proclaim Proposition G. An extension is Galois if and only if it is the splitting field of a family of separable polynomials.

\proof If an extension is the splitting field of a family of separable polynomials then it is clearly normal and separablility comes from the fact that $L/K$ is generated by the roots of separable polynomials. Conversely, if $\alpha\in L$, then the minimal polynomial of $\alpha$ splits into linear factors in $L$ with no multiple roots, so the extension is both normal and separable.\slug

If $K\subseteq L\subseteq M$ is a tower of extensions, then $M/K$ may not be Galois even though $M/L$ and $L/K$ are. In particular, the normality condition may fail. The standard example is
$$\Q \subseteq \Q[\sqrt 2] \subseteq \Q[\root 4\of 2].$$
The extension $\Q[\root 4\of 2]$ is not normal, but the intermediate extensions are quadratic extensions and therefore Galois. However, it is true that $M/K$ is separable if and only if $M/L$ and $L/K$ are separable.

Earlier, for an extension $L/K$ we defined an intermediate subfield $L_\sep\subseteq L$ called the separable closure of $K$ in $L$. This is the set of elements in $L$ which are separable over $K$. Since $L_\sep/K$ is separable, $[L_\sep:K]$ is exactly the number of $K$-embeddings from $L_\sep\inj \bar K$. Since $\bar K$ contains the roots of every polynomial in $K[x]$, we can define the {\it separable closure} of $K$ to be $K^\sep = \bar K_\sep$. A field is called {\it perfect} if every irreducible polynomial over it is separable. We will prove that $\bar K_\sep = \bar K$ if and only if $K$ is perfect.

For a polynomial $f\in K[x]$, we define its {\it derivative} $f'\in K[x]$ in the the usual way. If $f(x) = a_nx^n + \cdots + a_1x + a_0$, then $f'(x) = na_nx^{n-1} + \cdots + a_1$.

\proclaim Lemma D. Let $K$ be a field. Then $f$ is separable if and only if $f$ and $f'$ have no common root in $\bar K$, which is true if and only if $\gcd(f,f') = 1$.

\proof Observe that $\gcd(f,f')\neq 1$ if and only if $f$ and $f'$ have a common irreducible factor and this is equivalent to them having a common root in $\bar K$. If $\alpha\in \bar K$ is a root of $f$, then $f = (x-\alpha)g$ for some $g$ and by the product rule, $f' = g + (x-\alpha)g'$ so if $\alpha$ is also a root of $f'$, then $g(\alpha) = 0$ and $f = (x-\alpha)^2h$ for some $h$. Conversely, if $f = (x-\alpha)^2h$ for some $h$, then $\alpha$ is a root of both $f$ and $f'$.\slug

If $K$ is a field and $f$ is irreducible, then $f$ is separable if and only if $f' \neq 0$. If the characteristic is zero, then $\deg(f') = \deg(f) - 1$, so $f' \neq 0$ is equivalent to $f$ being a nonconstant irreducible. Thus fields of characteristic zero are perfect. A field of characteristic $p$ is perfect if the Frobenius map that takes $x\mapsto x^p$ is surjective.

\section 9. Fixed Fields

Let $G$ be a semigroup and $L$ be a field. A {\it multiplicative character} is a map $\chi : G\to L$ which is multiplicative:
$$\chi(st) = \chi(s)\chi(t)$$

\parenproclaim Lemma D (Dedekind's lemma). Multiplicative characters on a semigroup are linearly independent, i.e.\ if $\chi_1,\ldots,\chi_n$ are distinct and there exist $\beta_i\in L$ such that
$$\sum_{i=1}^n \beta_i\chi_i = 0,$$
then all $\beta_i$ are 0.

\proof The proof is by induction. If $n=1$, then $\beta_1\chi_1(s) = 0$ for all $s$ implies that $\beta = 0$. Now assume the lemma holds for $n$ and assume that $\sum_{i=1}^{n+1} \beta_i\chi_i = 0$. Then for any $s,t\in G$, we infer from multiplicativity that
$$\sum_{i=1}^{n+1} \beta_i\chi_i(s)\chi_i(t) = 0.$$
On the other hand,
$$\Big(\sum_{i=1}^{n+1} \beta_i\chi_i\Big)\chi_{n+1}(t) = 0,$$
so by subtracting the two equations we get
$$\sum_{i=1}^n \Big(\beta_i\big(x_i(t) - \chi_{n+1}(t)\big)\Big) \chi_i(s) = 0$$
for all $s\in G$. By the induction hypothesis, $\chi_1,\ldots,\chi_n$ are independent over $L$, so for $i = 1,\ldots,n$
$$\beta_i\big(\chi_{i}(t) - \chi_{n+1}(t)\big) = 0$$
for all $t\in G$. Now, $\chi_i \neq \chi_{n+1}$ if and only if there exists $t\in G$, $\chi_i(t) \neq \chi_{n+1}(t)$, so $\beta_i = 0$ for all $i \leq n$. Then $\beta_{n+1}\chi_{n+1} = 0$ so $\beta_{n+1} = 0$.\slug

Dedekind's lemma tells us that if $L/K$ is a field extension, then the set of homomorphisms from $K$ to $L$ is linearly independent. This provides an upper bound on the number of homomorphisms if the dimension of the extension is finite.

Let $L$ be a field and $G\subseteq \Aut(L)$ be a subgroup. The {\it fixed field} of $G$ is the field
$$L^G = \{\alpha\in L : \theta(\alpha) = \alpha\ \hbox{for all}\ \theta\in G\}.$$
To give a simple proof of the main result of the section, we will require the primitive element theorem. The proof given is the treatment given by van der Waerden.

\parenproclaim Theorem P (Primitive element theorem). Let $L = K(\alpha_1,\ldots,\alpha_n)$ be a finite algebraic extension of $K$ and suppose that $\alpha_2,\ldots,\alpha_n$ are separable (it is not required that $\alpha_1$ be separable). Then there exists an element $\gamma$ such that $L = K(\gamma)$.

\proof We can suppose that $L$ is infinite, since otherwise $K$ is finite and we can let $\gamma$ be a primitive root of unity that generates $K^\times$. Moreover, it suffices to show the theorem for two elements $\alpha$ and $\beta$, with $\beta$ separable, since a simple induction will extend the result to arbitrary $n$. Let $f$ and $g$ be irreducible polynomials for which $\alpha$ and $\beta$ are roots, respectively. Let $\alpha_1,\ldots,\alpha_r$ be the distinct roots of $f$ and $\beta_1,\ldots,\beta_s$ be roots of $g$; let $\alpha =\alpha_1$ and $\beta = \beta_1$.

For $k\neq 1$ we have $\beta_k\neq \beta_1$ so the equation
$$\alpha_i + \beta_k x = \alpha_1 + \beta_1x$$
has at most one solution in $K$ for every $i$ and every $k\neq 1$. If we take $c\in K$ to be different from any of these solutions, we have
$$\alpha_i + c\beta_k \neq \alpha_1 + c\beta_1$$
for every $i$ and $k\neq 1$. Now we let
$$\gamma = \alpha_1 + c\beta_1 = \alpha + c\beta;$$
this is an element of $K(\alpha,\beta)$.

The element $\beta$ satisfies
$$g(\beta) = 0\qquad\hbox{and}\qquad f(\gamma - c\beta) = f(\alpha) = 0,$$
with coefficients in $K(\gamma)$. The polynomials $g(x)$ and $f(\gamma - cx)$ only have the root $\beta$ in common, since for $k\neq 1$, $i = 1,\ldots,r$, $a_i \neq \theta - c\beta_k$ and $f(\gamma - c\beta_k) \neq 0$. Since $\beta$ is separable, the polynomials $g(x)$ and $f(\gamma - cx)$ only have the factor $x-\beta$ in common and the coefficients of this factor must lie in $K(\gamma)$, so $\beta\in K(\gamma)$. The same thing can be shown for $\alpha$ from the identity $\alpha = \gamma - c\beta$. So $K(\gamma) = K(\alpha,\beta)$.\slug

The main theorem of this section links fixed fields to Galois extensions.

\parenproclaim Theorem A (Artin's fixed field theorem). Let $L$ be a field and let $G$ be a subgroup of $\Aut(L)$.
\medskip
\item{i)} If $G$ acts with finite orbits, then $L$ is a Galois extension of $L^G$.
\smallskip
\item{ii)} If $|G|$ is finite, then $[L:L^G] = |G|$ and $G$ is the Galois group $\Gal(L/L^G)$.
\medskip

\proof Let $\alpha \in L\setminus L^G$ and let $\{\alpha_1,\ldots,\alpha_n\}$ be the orbit under the $G$-action. Then
$$p(x) = \prod_{i=1}^n (x-\alpha_i)$$
is $G$-invariant and $p\in L^G[x]$ is separable. $L$ is the splitting field of $p$ so $L$ is a Galois extension of $L^G$.

To prove (ii), suppose that $G$ is finite and let $n=|G|$. Take $\alpha\in L\setminus L^G$ and since $|G\alpha|\leq n$ for any $\alpha\in L$, $[L^G(\alpha) : L^G] \leq n$. We use this in our proof that $[L:L^G] \leq n$. Take $\alpha\in L$ such that $[L^G(\alpha):L^G]$ is maximal and let $\beta\in L$. Then $L^G(x,y)$ is a finite extension. By the primitive element theorem, $L^G(\alpha,\beta) = L^G(\gamma)$ for some $\gamma\in L$. But by the maximality of $\alpha$, we have $[L^G(\alpha):L^G]\geq L^G(\gamma):L^G]$ so $L^G(\alpha) = L^G(\gamma)$. This means that $\beta\in L^G(\alpha)$ and since our choice of $\beta$ was arbitrary, $L = L^G(\alpha)$. In particular, $[L:L^G]\leq n$.

Now if $[L:L^G] < n$, then $L$ cannot have $L$ automorphisms over $L^G$. But $G$ is a subgroup of $\Aut(L/L^G)$ with $n$ elements. So $[L:L^G] = n$ and $G = \Aut(L/L^G)$.\slug

This theorem tells us that finite groups give field extensions, in the following sense.

\proclaim Corollary G. If $G$ is a finite group, then there exists an extension $L/K$ with $G = \Aut(L/L^G)$.

\proof By the fixed field theorem, it suffices to represent $G$ as a group of automorphisms over some field $L$. Let $\{x_s\}_{s\in G}$ be a set of $|G|$ indeterminate variables over some field $F$. Then $G = \{e,s,t,\ldots\}$ acts on the polynomial ring $F[x_e,x_s,x_t,\ldots]$ by acting on the variables:
$$sx_t = x_{st}$$
The action extends to monomials as well; for a subset $T\subseteq G$,
$$s\Big(\prod_{t\in T} x_t\Big) = \prod_{t\in T} x_{st}$$
and the action of $G$ on $F[x_s : s\in G]$ is by ring automorphisms. We can extend the action to the fraction field $L = F(x_s : s\in G)$ and the result is a faithful action by field automorphisms. Thus $G\subseteq \Aut(L)$ and $L$ has degree $|G|$ over $L^G$.\slug

A well-known open problem is the following: {\sl Given a finite group $G$, is it possible to to find an extension $L/\Q$ such that $G = \Aut(L/\Q)$ and $\Q = L^G$? Equivalently, is $G$ the Galois group of a polynomial with rational coefficients?} For example, the group of quaternions $Q_8$ is the Galois group of the polynomial
$$f = x^8 - 72x^6 + 180x^4 - 144x^2 + 36$$
and the group ${\rm PSL}_2(7)$ of order 168 (the second non-abelian simple group) is the Galois group of
$$x^7 - 56x^6 + 609x^5 + 1190x^4 + 6536x^3 + 4536x^2 - 6804x - 5832 - tx^3(x + 1),$$
where the parameter $t$ can be any integer $t\equiv 1\mod{31}$.

\proclaim Proposition S. If $f\in \Q[x]$ is irreducible of prime degree $p$ with exactly two roots in $\C\setminus \R$, then the Galois group of $f$ is $S_p$.

\proof Complex conjugation is a non-trivial Galois automorphism, which acts as a transposition since there are two non-real roots. We just need to show that there exists a $p$-cycle. But the number $p$ of roots of $f$ divides the order of the Galois group, so by Cauchy's theorem we have an element of order $p$. This must be a $p$-cycle and we have generated $S_p$.\slug

\section 10. The Galois Correspondence

If $L/K$ is a Galois extension, then $\alpha\in L\setminus K$ must be displaced by some automorphism. The converse is also true, giving us the following theorem.

\parenproclaim Theorem F (Galois' fixed field theorem). Let $L/K$ be an algebraic extension with automorphism group $G = \Aut_K(L)$. Then $L/K$ is a Galois extension if and only if $K = L^G$.

\proof For the forward direction, let $\alpha\in L^G$ have minimal polynomial $f$ and splitting field $M$. Because $M$ is normal, we assume that $M\subseteq L$. If $\beta\in M$ is another root of $f$, then we can find $\theta\in \Aut_K(M)$ such that $\theta(\alpha) = \beta$. Extend $\theta$ to an element in $\Aut_K(L)$. Since $\alpha\in L^G$, we have $\beta = \theta(\alpha)  = \alpha$, i.e.\ $f$ has only one root. Since $L$ is separable, $f$ has degree 1, so $\alpha\in K$.

For the converse, assume that $K = L^G$. Let $\alpha\in L$ have minimal polynomial $f$ with set of roots $R\subseteq \bar L$, and let
$$g = \prod_{\beta \in R\cap L} (x - \beta).$$
Note that $g$ is separable and invariant under every $\theta\in \Aut_K(L)$, since $\theta(R\cap L) = R\cap L$. This implies that the coefficients of $g$ are fixed by $G = \Aut_K(L)$. Since $L^G = K$, $g\in K[x]$ and since $g(\alpha) = 0$, it divides the minimal polynomial. Thus $f =g$ and all the roots of $f$ are in $L$, proving normality. Lastly, $f$ is separable because $g$ is.\slug

The Galois correspondence was used by Galois to compute Galois groups of polynomial equations. It comprises the following results:
\medskip
\item{i)} Equations that are solvable by radicals have solvable Galois groups.
\smallskip
\item{ii)} Conversely, if $\Gal(f)$ is solvable, then the roots of $f$ can be expressed as radicals.
\smallskip
\item{iii)} $A_5$ is the smallest group that is not solvable (so every equation of degree $\leq 4$ can be solved by radicals).
\smallskip
\item{iv)} $A_5$ is the Galois group of a polynomial of degree 5 with rational coefficients, which is therefore not solvable by radicals.
\medskip
If $L/K$ is a Galois extension, then $L/M$ is Galois for any intermediate extension $K\subseteq M\subseteq L$.

\parenproclaim Theorem C (Classical Galois correspondence). Let $L/K$ be a finite Galois extension. Then there is a bijection $\phi$ from subgroups of $\Gal(L/K)$ and intermediate fields in $L/K$. This is explicitly given by $\phi(H) = L^H$ for $H\subseteq \Gal(L/K)$ and $\phi^{-1}(M) = \Gal(L/M)$ for an intermediate extension $K\subseteq M\subseteq L$.

\proof From Artin's fixed field theorem and Theorem F we know that $|G| = [L:K]$ and that the map $\phi$ is injective. Let $M$ be an intermediate extension and let $H = \Gal(L/M)$. Observe that since $L/M$ is Galois, Theorem F asserts that $M = L^H$. Thus $\phi^{-1}$ is injective and we have a bijection.\slug

The correspondence is order-reversing in that the bigger the subgroup $H$, the smaller the fixed field $L^H$. Another corollary is that a Galois extension can only have finitely many intermediate subfields and that the number of subfields is at most the number of subgroups in the Galois group.

\proclaim Proposition N. Let $L/K$ be a Galois extension and $M$ be an intermediate field. The following are equivalent:
\medskip
\item{i)} $\Gal(L/M)$ is a normal subgroup of $\Gal(L/K)$.
\smallskip
\item{ii)} If $s\in \Gal(L/K)$, then $s(M) = M$.
\smallskip
\item{iii)} $M/K$ is a normal extension.
\smallskip
\item{iv)} $K = M^H$ where $H = \Aut_K(M)$.
\smallskip
\item{v)} If $M/K$ is a finite extension, then $[M:K] = |\Aut_K(M)|$.
\smallskip
\item{vi)} The map $\Gal(L/K)\to \Gal(M/K)$ that maps $s\mapsto s_{|M}$ is surjective with kernel $\Gal(L/M)$.\noskipslug
\medskip

Let $L/K$ be a Galois extension. Let $K\subseteq M,N\subseteq L$ be intermediate subfields. The {\it compositum} $MN$ of $M$ and $N$ is the subfield of $L$ that they generate, i.e.\ the smallest subfield of $L$ containing them. It can be shown that
$$MN = \bigg\{\sum_{i=1}^m \sum_{j=1}^n \alpha_i\beta_j : \alpha_1,\ldots,\alpha_m \in M, \beta_1,\ldots,\beta_n\in N\bigg\}.$$

\proclaim Proposition I. If $M/K$ and $N/K$ are Galois, then so is the compositum $MN$.

\proof It suffices to show normality. Suppose that $M$ and $N$ are finite normal extensions. Write $M = K(R_1)$ and $N=K(R_2)$ as splitting fields of two polynomials $f_1$ and $f_2$ with respective root sets $R_1$ and $R_2$. Then the splitting field of the product $f = f_1f_2$ is $K(R_1\cup R_2)$, which is the compositum $MN = K(R_1)K(R_2)$. If $M$ and $N$ are infinite, we can reuse the argument by writing $M$ and $N$ as splitting fields of families $F_1$ and $F_2$ of polynomials.\slug

The Galois correspondence interchanges intersection and compositum, as shown by the following proposition.

\proclaim Proposition S. Let $L/K$ be a finite Galois extension and let $G$ and $H$ be subgroups of $\Gal(L/K)$. Then
\medskip
\item{i)} $L^G \cap L^H = L^{\langle G,H \rangle}$;
\smallskip
\item{ii)} $L^GL^H = L^{G\cap H}$.
\medskip

\proof This is a corollary of Theorem C. Because $\langle G,H\rangle$ is the smallest subgroup containing both $G$ and $H$, it corresponds to the largest subfield $L^G\cap L^H$ of both $L^G$ and $L^H$. Likewise, the compositum $L^GL^H$ is the smallest subfield containing both $L^G$ and $L^H$, so it corresponds to the largest subgroup $G\cap H$ of both $G$ and $H$.\slug

If $M$ and $N$ are Galois then $MN$ is Galois and we have a homomorphism
$$f : \Gal(MN/K) \to \Gal(M/K) \times \Gal(N/K)$$
that maps $\sigma\mapsto (\sigma_{|M},\sigma_{|N})$. It can be shown that $\ker f = \{e\}$ and $\Im f$ consists of pairs $(s,t)$ for which $s_{|M\cap N} = t_{|M\cap N}$. In terms of polynomials, if $f = f_1f_2$, the Galois group of $f$ is the subgroup of $\Gal(f_1)\times \Gal_(f_2)$ consisting of pairs of permutations that coincide on $K(R_1)\cap K(R_2)$, where $R_1$ and $R_2$ are the root sets of $f_1$ and $f_2$.

Thus, if $L/K$ is a Galois extension such that $L = MN$ for normal extensions $M/K$ and $N/K$, and furthermore $M\cap N = K$, then $\Gal(L/K)$ can be expresed as the direct product
$$\Gal(L/K)\cong \Gal(M/K) \times \Gal(N/K).$$


\bye
