% Marcel Goh CV

% From cwebmac: Date and time
\def\today{\ifcase\month\or
  January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or December\fi
  \space\number\day, \number\year}
\newcount\twodigits
\def\hours{\twodigits=\time \divide\twodigits by 60 \printtwodigits
  \multiply\twodigits by-60 \advance\twodigits by\time :\printtwodigits}
\def\gobbleone1{}
\def\printtwodigits{\advance\twodigits100
  \expandafter\gobbleone\number\twodigits
  \advance\twodigits-100 }

\font\ninesc=cmcsc9
\font\eightrm=cmr8
\font\elevensc=cmcsc10 at 11pt  % For section headers
\font\bigbold=cmb10 at 12pt  % Name
\font\tenb=cmb10
\font\mc=cmr9

\let\tenbf=\tenb

% For publication lists (adapted from Knuth's algorithms)
\newdimen\itemindent
\newif\ifitempar \itempartrue
\def\itemindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\itemindent=\wd0\relax}
\def\pubbegin #1{\itemindentset{#11}} % when all have 1 digit
\def\ppubbegin #1{\itemindentset{#111}} % when 10 or more publications
\def\pubitem#1.{\ifitempar\smallskip\noindent\else\itempartrue
  \hskip-\parindent\fi
  \hbox to\itemindent{\bf\hfil #1.\kern.25em}%
  \hangindent=\itemindent\hangafter=1\ignorespaces}

\def\up#1{\leavevmode \raise.16ex\hbox{#1}}
\def\sectheader#1{{\bigskip\elevensc #1}\smallskip\hrule\medskip}
\def\leftright#1#2{\hbox to\hsize{{#1}\hfill{#2}}\par}
\def\starleftright#1#2{\hbox to\hsize{\llap{*}{#1}\hfil{#2}}\par}
\def\CEE{{\mc C}}
\def\UNIX{{\mc UNIX}}
\def\thing{\item{$\bullet$}}

\parindent=0pt  % No indentation

\newcount\papercount
\papercount=1
\def\papitem{\pubitem P\the\papercount. \global\advance\papercount by -1}
\newcount\submittedcount
\submittedcount=1
\def\subitem{\pubitem S\the\submittedcount. \global\advance\submittedcount by -1}
\newcount\reportcount
\reportcount=1
\def\repitem{\pubitem R\the\reportcount. \global\advance\reportcount by -1}

\def\datethis{\eightrm \today}

% =========== START ============ %

% \footline{\phantom{\datethis}\hss\folio\hss\datethis}  % \phantom for centering

\leftright{\bigbold Marcel Kieren Goh
\ifx\Umathchar\undefined\else% Chinese characters do not appear when using pdfTeX
\font\chinese="AR PL UKai HK" at 12pt%
\font\twelverm=cmr12%
\twelverm\up({\chinese 吳麒仁}\up)%
\fi}{\datethis}
\smallskip

\sectheader{Contact Information}

\leftright{Mobile phone: +1\thinspace\up(825\up)\thinspace 440-0681}{2-9602 74 Avenue NW}
\leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}

\sectheader{Languages}

Fluent: English, French. Proficient: Mandarin, Italian.

\sectheader{Education}

\leftright{\bf M.Sc.\ Mathematics}{\bf May 2021 -- present}
\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
\smallskip
Supervisor: Luc Devroye. CGPA: 4.0/4.
Thesis: ``Structural properties of conditional Galton--Watson trees.''
\medbreak
\leftright{\bf B.Sc.\ Joint Honours Mathematics and Computer Science}{\bf September 2017 -- April 2021}
\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
\smallskip
Minor: Linguistics. CGPA: 3.84/4.
\medbreak
\leftright{\bf Exchange semester}{\bf February -- June 2019}
\leftright{\sl Faculty of Mathematics and Physics, Charles University}{\sl Prague, Czech Republic}
\smallskip
Grade: 1 on all courses (highest score attainable).


\sectheader{Work and Volunteer Experience {\rm($*$ indicates a paid position)}}

\starleftright{\bf Visiting teaching assistant}{\bf July 2022}
\leftright{\bf Bocconi University}{\sl Como, CO, Italy}
\smallskip
Teaching assistant for a two-week summer school on random structures and combinatorial statistics,
which was
organised by Bocconi University in collaboration with Oxford University
and Imperial College London. Led problem sessions in the afternoons;
around forty graduate-level students were in attendance.
\medbreak

\leftright{\bf Mentor}{\bf January -- April 2022}
\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
\smallskip
Met with two undergraduate students as part of the Directed Reading Program
on a weekly basis to give them an introduction to research-level
mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
triples problem, as well as topics related to Khovanskii's theorem.
\medbreak

\starleftright{\bf Teaching assistant}{\bf September -- December 2021}
\leftright{\sl School of Computer Science, McGill University}{\sl Montr\'eal, QC, Canada}
\smallskip
Teaching assistant for COMP 690, a graduate-level course on the probabilistic analysis of algorithms.
Held office hours twice a week and responsible for grading of assignments.
\medbreak

\leftright{\bf First responder}{\bf October 2017 -- September 2021}
\leftright{\sl McGill Student Emergency Response Team}{\sl Montr\'eal, QC, Canada}
\smallskip
On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
first-aid skills up-to-date. Most recent first responder certification: September 2020.
\medbreak

\starleftright{\bf Grader}{\bf September 2019 -- April 2021}
\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
\smallskip
Grading of assignments in the following courses:
\begingroup\parindent=10pt
\smallskip
\thing Winter 2021: MATH 457 Honours Algebra 4
\smallskip
\thing Fall 2020: MATH 323 Probability, MATH 456 Honours Algebra 3
\smallskip
\thing Winter 2020: MATH 240 Discrete Structures
\smallskip
\thing Fall 2019: MATH 235 Algebra 1
\endgroup
\medbreak

\starleftright{\bf Helpdesk tutor}{\bf September 2018 -- April 2021}
\leftright{\sl Computer Science Undergraduate Society Helpdesk}{\sl Montr\'eal, QC, Canada}
\smallskip
Hold twice-weekly office hours to tutor students in a variety of undergraduate courses.
Topics covered include elementary data structures and algorithms, command-line scripting,
and functional programming. Recipient of the Tomlinson Engagement Award for Mentoring.
\medbreak

\leftright{\bf Vice President, Academic}{\bf May 2019 -- January 2020}
\leftright{\sl Society of Undergraduate Mathematics Students}{\sl Montr\'eal, QC, Canada}
\smallskip
Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
mathematics faculty. Duties included representing the student body at department meetings,
organising midterm and final review sessions,
and helping students with
academic concerns.
\medbreak

\starleftright{\bf Painter}{\bf May -- August 2018}
\leftright{\sl Bakir Contracting Corp.}{\sl Edmonton, AB, Canada}
\smallskip
Exterior painting (siding, decks, fences, trim, etc.) for residential clients.
\medbreak

\starleftright{\bf Infantryman}{\bf August 2015 -- August 2017}
\leftright{\sl Singapore Armed Forces}{\sl Singapore}
\smallskip
Held appointment of machine-gun team commander in the 3rd Battalion, Singapore Infantry Regiment.
Led a six-person team consisting of a medic, signaller, sensor, and two-machine gunners within a rifle platoon.
\medbreak

\starleftright{\bf Tutor}{\bf September 2014 -- June 2015}
\leftright{\sl \'Ecole Secondaire Beaumont Composite High School}{\sl Beaumont, AB, Canada}
\smallskip
Tutored various students in grades 4 through 11 in chemistry, physics, math, and French.
\medbreak

\starleftright{\bf Summer camp counsellor}{\bf June -- August 2014}
\leftright{\sl YoWoChAs Outdoor Education Centre}{\sl Fallis, AB, Canada}
\smallskip
Led children aged 4--15 through various activities (e.g. archery, canoeing, zipline) at a sleepaway camp.
\medbreak

\sectheader{Awards and Funding}

\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
\smallskip
\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
\smallskip
\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}

\sectheader{Papers}

\begingroup\frenchspacing
\papercount=5
\pubbegin P
\papitem (with Jad Hamdan and Jonah Saks)
The lattice of arithmetic progressions. To appear in {\sl Australasian Journal of Combinatorics}, 15~pp.
[arXiv:2106.05949]
\papitem (with Luc Devroye and Rosie Y. Zhao)
On the peel number and the leaf-height of a Galton--Watson tree.
To appear in {\sl Combinatorics, Probability and Computing}, 20~pp. [arXiv:2106.14389]
\papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
and Theoretical Computer Science} {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
\papitem (with Anna M. Brandenberger and Luc Devroye)
Root estimation in Galton--Watson trees. {\sl Random Structures and Algorithms} {\bf 61},3 (2022), 520--542.
[arXiv:2007.05681]
\papitem (with Rosie Y. Zhao)
Arithmetic subsequences in a random ordering of an additive set.
{\sl Integers: Electronic Journal of Combinatorial Number Theory} {\bf 21} (2021), \#A89, 19~pp.
[arXiv:2012.12339]

\sectheader{Submitted Papers}

\begingroup\parindent=10pt
\thing (with Jonah Saks)
Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
\endgroup
\endgroup%end the frenchspacing


\bye

